---
title: "Data censoring in social sciences"
author: "Modified by Yihan Du based on past code by Xiaotang, Bikramjit, and Karthik"
date: "Term 5, 2026"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

# Data preparation

First, we prepare the working environment, load the survival package and `extramarital.csv` data.
```{r}
rm(list=ls()) # Remove all variables from the R environment to create a fresh start
# setwd("...")  # Setup the working directory
```

### Installation

Next we install the `survival` package, which includes functions for running the Tobit model.

```{r}
if(!require(survival)){
  install.packages("survival")}
library(survival)
```  

Next we read the "extramarital.csv" file.
```{r}
exm <- read.csv("extramarital.csv")
str(exm)
```

The dataset consists of 6366 observations of 7 variables:

 - `marriage_rating`: 5 = very good, 1 = very poor
 - `age`: 17.5 = under 20, 22 = 20 to 24, ..., 42 = 40 or over 
 - `yrs_married`: 0.5 = less than 1 year, 2.5 = 1 to 4 years ...
 - `religiosity`: 4 = strong, ..., 1 = no
 - `education`: level of education, 9 = grade school, 12 = high school, ... 
 - `occupation`: 6 = professional with advanced degree, 5 = managerial, administration, ... 1 = student
 - `time_in_affairs`: amount of time spent in extra-marital affairs

The dataset was collected in a survey by Redbook and analyzed by Ray C. Fair (Yale).

# Modeling and Analysis

##  Modelling 
  Let's take a better look at the data
```{r}  
table(exm$time_in_affairs>0)
```
We observe that about 32% of people in the dataset are having an affair. 
<!-- # FALSE  TRUE  -->
<!-- # 4313  2053  -->

Let's look at the distribution of `time_in_affairs` (with 100 bins).
```{r}
hist(exm$time_in_affairs,100)
```

Note that in this dataset there are a large number of zeros for the affair variable. If someone is not having an affair, a small decrease in one of the predictors may not change his/her  chances of having an affair in comparison to a person having an affair. Hence, one needs to account for possibly two types of people: those who are unlikely to be moved away from 0 by modest changes in the predictors, and those who are not 0 and likely to change by modest change in predictors.
To do this, we use the *Tobit* model.

## Training and test dataset

We create a training and a test dataset first.
```{r}
set.seed(100)
spl <- sample(nrow(exm),0.7*nrow(exm)) # Randomly select 70% of rows
train <- exm[spl,]
test <- exm[-spl,] # Remaining rows not in spl
str(train)
str(test)
```

## Tobit model
Now we fit the  Tobit model.
```{r}
?Surv     # Create a survival object, usually used as a response variable in a model formula.
?survreg  # Fit a parametric survival regression model.
model1 <- survreg(Surv(time_in_affairs,time_in_affairs>0,type="left")~.,data=train,dist="gaussian")
## time_in_affairs: the observed variable (e.g., duration of an affair).
## time_in_affairs > 0: this is a logical indicator — TRUE means the value is not censored (we know the exact value), and FALSE means it's censored (we only know it's less than or equal to the observed value).
## type = "left": this is left-censoring.
```

This fits a Tobit model where the data are left censored at zero. The error terms are assumed to be Gaussian (normally distributed). Let us look at the model summary.
```{r}
 summary(model1)
```

Now, let's look at the estimated parameters ($\beta$-coefficients. )
 
 - `marriage_rating`: negative, meaning that as the marriage rating increases, the chances/length of an affair decrease.
 - `occupation`: positive, meaning that there are greater chances of an affair with a better occupation.
 - `yrs_married`: positive
 - `religiosity`: negative

Also, note that all coefficients are fairly significant. The model has a log likelihood of -5436.9.

### Prediction
To check how good our model fits, we predict on the test data 

```{r}
predict1 <- predict(model1,newdata=test) # Prediction provides the latent predicted values of the variables (positive and negative)
# plot(predict1)
```

## Benchmark model

We also use a benchmark model, namely *linear regression* to compare.
```{r}
model2 <- lm(time_in_affairs~.,data=train)
# summary(model2)
predict2 <- predict(model2,newdata=test)
# plot(predict2)
```


## Model comparison

Now let's compare the two models
```{r}
table(predict1 <= 0, test$time_in_affairs==0)
```
<!-- #       FALSE TRUE -->
<!-- # FALSE    129   48 -->
<!-- # TRUE    473 1260 -->
```{r}
table(predict2 <= 0, test$time_in_affairs==0)
```
<!-- #       FALSE  TRUE -->
<!-- # FALSE   571 1184 -->
<!-- # TRUE    31   124 -->
<!-- # -->

Clearly, the Tobit model has better accuracy in terms of accounting for  the large number of individuals for whom the extra-marital affairs variable was 0.

Note that here we simply classify observations according to whether the predicted latent mean $\beta_0+\sum_{j=1}^{p}\beta_j x_{ij}$ is above or below zero. 
A more principled approach is to compute the probabilities
\[
\mathbb P(Y_i>0)=\mathbb P(Y_i^*>0),
\qquad
\mathbb P(Y_i=0)=\mathbb P(Y_i^*\le 0),
\]
since the Tobit model specifies a full probability distribution for the latent variable $Y_i^*=\beta_0+\sum_{j=1}^{p}\beta_j x_{ij}+\epsilon_i$ rather than only its predicted mean.


##### End of Data censoring for social sciences analyses.

